Does Anybody Really Know What Time It Is?\\ Automating the Extraction of Date Scalars

### 2015-07-16 futurework.tex
Edits from Rick Cole's paper pass.

While we are satisfied with the results so far, we found a number of column that we were not able to process with these techniques. Many columns contained multiple date formats, and we would like to recognize this situation and generate predicate-based calculations to increase the accuracy of the results and thereby make the experience even more seamless for the user. Other columns contain date ranges, which we would like to handle by generating multiple calculations, possibly by combining regular expressions with date parsing.

>>>>

While we are satisfied with the results so far, we found a number of columns that we were not able to process with these techniques. Many columns contained multiple date formats. We would like to recognize this situation and generate predicate-based calculations to increase the accuracy of the results and thereby make the experience even more seamless for the user. Other columns contain date ranges, which we would like to handle by generating multiple calculations, possibly by combining regular expressions with date parsing.

### 2015-07-16 intro.tex
Edits from Rick Cole's paper pass.

\subsubsection{Function Libraries}
While the existing standards for query languages are helpful for defining a core set of row-level functions, there are many useful functions beyond the standards that are only explicitly supported by a subset of RDBMSes, often with different names. And even when a database does not provide an implementation of a particular function, it is usually possible to generate it as an inline combination of existing functionality. The calculation languages of visualization systems can thus be designed to provide a \textit{lingua franca} that tames this Babel of dialects by providing a common syntax for as many of these functions as possible.

@@

\subsubsection{Parsing}
This parsing of date representations is one of the oldest data preparation problems we have observed. A common version of this problem is how to convert columns of integers of the form \texttt{yyyyMMdd} to date scalars. Na\"{i}ve users often solve this problem by converting the integer to a string and performing some locale-dependent string operations before casting the string back to a date. Unfortunately, this approach has a number of problems:
\begin{itemize}
\item String operations are notoriously slow compared to scalar operations (typically 10-100x slower in modern RDBMSes)
\item The parsing code was hard to understand and maintain because it used a verbose, general-purpose string-handling syntax instead of a specialized domain language.
\item Default parsing of date formats is locale-dependent, and may not work when the analysis is shared across an international organization (\textit{e.g} between the US and European offices)
\item The parsing code is hard to understand and maintain because it uses a verbose, general-purpose string-handling syntax instead of a specialized domain language.
\end{itemize}

>>>>

\subsubsection{Function Libraries}
While the existing standards for query languages are helpful for defining a core set of row-level functions, there are many useful functions beyond the standards that are only explicitly supported by a subset of RDBMSes, often with different names. Even when a database does not provide an implementation of a particular function, it is usually possible to generate it as an inline combination of existing functionality. The calculation languages of visualization systems can thus be designed to provide a \textit{lingua franca} that tames this Babel of dialects by providing a common syntax for as many of these functions as possible.

@@

\subsubsection{Parsing}
This parsing of date representations is one of the oldest data preparation problems we have observed. A common version of this problem is how to convert columns of integers of the form \texttt{yyyyMMdd} to date scalars. Na\"{i}ve users often solve this problem by converting the integer to a string and performing some locale-dependent string operations before casting the string back to a date. Unfortunately, this approach has a number of problems:
\begin{itemize}
\item String operations are notoriously slow compared to scalar operations (typically 10-100x slower in modern RDBMSes)
\item Default parsing of date formats is locale-dependent, and may not work when the analysis is shared across an international organization (\textit{e.g} between the US and European offices)
\item The parsing code is hard to understand and maintain because it uses a verbose, general-purpose string-handling syntax instead of a specialized domain language.
\end{itemize}

### 2015-07-16 nlp.tex
Removed the spatial reference.
Edits from Rick Cole's paper pass.

We define a \textit{EBNF} grammar for identifying date-time strings below:

\begin{grammar}
<TimeGrammar> ::= <Hours> ':' <Minutes> ':' <Seconds> <TimeZone> <AMPM> (for 12-hour formats)

@@

The date-time grammar also includes a large number of syntactically correct but semantically invalid

date-time expressions. While we have added range restrictions to symbols such as \texttt{Hour} (1--12 for 12-hour format and 1--24 for 24-hour format), \texttt{Days} (1--7), \texttt{Month} (1--12), there are special cases that need to be accounted for. For example, there is no  \texttt{`November 31, 2015'}, \texttt{`February 29, 2013'}, or \texttt{`Sunday, May  5, 1965'}. November only as 30 days in any year; 2013 was not a leap year; and May 5, 1965 was a Wednesday.

@@

\item LeapYear Restriction for \texttt{February}:
While the above restriction applies to all months barring \texttt{February}, we also apply a constraint to the number of days for February, based on whether the year is leap year or not. For this, we define a new symbol in the grammar called \texttt{LeapYear}. If an expression containing the month \texttt{`February'} or any such variant (\textit{e.g.} \texttt{`Feb'}, \texttt{`2'}) with the day \texttt{`29'} and a year, would need to resolve the \texttt{Year} symbol to be a \texttt{LeapYear}, defined as:

>>>>

We define a \textit{EBNF} grammar for identifying date-time strings as follows:

\begin{grammar}
<TimeGrammar> ::= <Hours> ':' <Minutes> ':' <Seconds> <TimeZone> <AMPM> (for 12-hour formats);

@@

The date-time grammar also includes a large number of syntactically correct but semantically invalid date-time expressions. While we have added range restrictions to symbols such as \texttt{Hour} (1--12 for 12-hour format and 1--24 for 24-hour format), \texttt{Days} (1--7), \texttt{Month} (1--12), there are special cases that need to be accounted for. For example, there is no  \texttt{`November 31, 2015'}, \texttt{`February 29, 2013'}, or \texttt{`Sunday, May  5, 1965'}. November only as 30 days in any year; 2013 was not a leap year; and May 5, 1965 was a Wednesday.

@@

\item LeapYear Restriction for \texttt{February}:
While the above restriction applies to all months barring \texttt{February}, we also apply a constraint to the number of days for \texttt{February}, based on whether the year is leap year or not. For this, we define a new symbol in the grammar called \texttt{LeapYear}. If an expression containing the month \texttt{`February'} or any such variant (\textit{e.g.} \texttt{`Feb'}, \texttt{`2'}) with the day \texttt{`29'} and a year, would need to resolve the \texttt{Year} symbol to be a \texttt{LeapYear}, defined as:

### 2015-07-16 testing.tex
Edits from Rick Cole's paper pass.

\subsubsection{Natural Language Processing}
We implemented the grammar, training and parsing in Python 3.4.1 using the Natural Language Toolkit libraries~\cite{nltk} on a quad core Dell T7600 running Windows 7. The complexity of the CYK parsing algorithm is $O(n^{3}|G|)$, where $n$ is the length of the input and $|G|$ is the size of the grammar~\cite{Younger67}. The PCFG grammar uses a set of $22$ phrase-level non-terminals and $30$ terminals to classify the various constituents in the corpus. Since the CYK algorithm finds the maximum likelihood solution, there are no search errors (rather probability $p = 0$, and a modification in the grammar is required to improve accuracy.

>>>>

\subsubsection{Natural Language Processing}
We implemented the grammar, training and parsing in Python 3.4.1 using the Natural Language Toolkit libraries~\cite{nltk} on a quad core Dell T7600 running Windows 7. The complexity of the CYK parsing algorithm is $O(n^{3}|G|)$, where $n$ is the length of the input and $|G|$ is the size of the grammar~\cite{Younger67}. The PCFG grammar uses a set of $22$ phrase-level non-terminals and $30$ terminals to classify the various constituents in the corpus. Because the CYK algorithm finds the maximum likelihood solution, there are no search errors (rather probability $p = 0$, and a modification in the grammar is required to improve accuracy.
